---
title: "Week 07 - Regression"
subtitle: "Model Fit, LM for Randomised Experiments, Multiple Predictors, Merging Data Sets in R<html><div style='float:left'></div><hr color='#EB811B' size=1px width=800px></html>"
author: "Danilo Freire"
date: "8th March 2019"
output: 
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    nature: 
      beforeInit: "macros.js"
      highlightStyle: github
      ratio: 4:3
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
r <- getOption("repos")
r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
options(repos = r)
set.seed(12345)
```

<style>

.remark-slide-number {
  position: inherit;
}

.remark-slide-number .progress-bar-container {
  position: absolute;
  bottom: 0;
  height: 6px;
  display: block;
  left: 0;
  right: 0;
}

.remark-slide-number .progress-bar {
  height: 100%;
  background-color: #EB811B;
}

.orange {
  color: #EB811B;
}
</style>

# Today's Agenda

.font150[
* Review of least squares

* How to assess model fit with $R^2$

* Analyse randomised experiments with `lm()`

* Merge data sets with `merge(x, y, by = )`
]
---

# Review

.font150[
* Linear regression minimises the squared distance of the points

* A few things to remember:

* .orange[Regression equation]: $Y = \alpha + \beta X + \epsilon$
  - $\alpha$ is the intercept (value of $Y$ when $X = 0$)
  - $\beta$ is the slope (increase/decrease in $Y$ when $X$ varies by one unit)
  - $\epsilon$ is the error term (difference from points to fitted line; mean zero)
]
---

# Linear Models

.font120[
```{r lm01,fig.align="center",tidy=F,warning=F,message=F,cache=T}
bivariate <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/prediction/bivariate_data.csv")
bivariate <- subset(bivariate, Year == 2010)
model.cm <- lm(Child.Mortality ~ log(GDP), data = bivariate)
model.cm
```
* $Y = 276.58 - 26.13X + \epsilon$
]
---

# R-Squared

.font140[
* How well does our regression line fit the data?

* $R^2$, _coefficient of determination_

$$R^2 = 1 - \frac{\textsf{SSR}}{\textsf{Total sum of squares (TSS)}}  \ = \ 1 - \frac{\sum_{i=1}^n \hat\epsilon_i^2}{\sum_{i=1}^n (Y_i - \overline{Y})^2}$$
* $R^2$ is also defined as the _explained variance_ in Y

* Goes from 0 to 1

* We can use `summary(model)` to see it
]
---

# Example

.font110[
```{r rsq2,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=18}
summary(model1)
```
]
---

# Caveats

.font150[
* $R^2$ cannot be used to compare models estimated with different samples

* $R^2$ .orange[does not] indicate whether:
  - _the independent variable causes the dependent variable_
  - the model is correctly specified 
  - the model suffers from ommitted variable bias
  - there are enough data points to make a valid conclusion
]
---

# Linear Models and RCTs

.font150[
* .orange[When the data come from a randomised experiment,] model parameters have a causal interpretation

* Treatment status as the independent variable (0 or 1)
  - 0 indicates control group
  - 1 indicates treatment group
]
--
.font150[
* $Y = \alpha + \beta \times treatment + \epsilon$
]
--
.font150[
* What is the interpretation of $\alpha$ here?

* What is the interpretation of $\beta$ here?
]
---

# Women as Policy Makers

.font120[
* Do women promote different policies than men?

* Observational studies: compare policies adopted by female politicians with those adopted by male politicians

* Randomised natural experiment:
  - one third of village council heads reserved for women
  - assigned at the level of Gram Panchayat (village council) since mid-1990s
  - each GP has multiple villages

* Hypothesis: female politicians represent the interests of female voters

* Female voters complain about drinking water while male voters complain about irrigation
]
---

# Data

| Name         | Description                                                                                                                |
| :----------- | :------------------------------------------------------------------------------------------------------------------------- |
| `GP`         | An identifier for the Gram Panchayat (village council)                                                                     |
| `village`    | identifier for each village                                                                                                |
| `reserved`   | binary variable indicating whether the GP was reserved for women leaders or not                                            |
| `female`     | binary variable indicating whether the GP had a female leader or not                                                       |
| `irrigation` | variable measuring the number of new or repaired irrigation facilities in the village since the reserve policy started     |
| `water`      | variable measuring the number of new or repaired drinking-water facilities in the village since the reserve policy started |

```{r data,fig.align="center",tidy=F,warning=F,message=F,cache=T}
women <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/prediction/women.csv")
names(women)
```
---

# Models

.font110[
* Does the reservation policy increase female politicians?

```{r m1,fig.align="center",tidy=F,warning=F,message=F,cache=T}
tapply(women$female, women$reserved, mean)
```

* Does it change the policy outcomes?
```{r m2,fig.align="center",tidy=F,warning=F,message=F,cache=T}
# drinking-water facilities
tapply(women$water, women$reserved, mean)

# irrigation facilities
tapply(women$irrigation, women$reserved, mean)
```
]
---

# Slope Coefficient = Difference-in-Means

.font120[
```{r women03,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=2}
tapply(women$water, women$reserved, mean)
```
```{r women04,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=1}
mean(women$water[women$reserved == 1]) - mean(women$water[women$reserved == 0])
```
```{r women05,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=7}
lm(water ~ reserved, data = women)
```
]
---

# Resume Experiment Revisited

.font100[
```{r soc01,fig.align="center",tidy=F,warning=F,message=F,cache=T}
resume <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/causality/resume.csv")
model1 <- lm(call ~ race, data = resume)
model1
tapply(resume$call, resume$race, mean)
# baseline (black) + the effect of having a white-sounding name
0.06448 + 0.03203
```
]
---

# Call Rates and Gender

.font110[
* Let's add a gender indicator

```{r soc02,fig.align="center",tidy=F,warning=F,message=F,cache=T}
model2 <- lm(call ~ sex, data = resume)
model2
tapply(resume$call, resume$sex, mean)
```
]
---

# Race + Gender

.font120[
```{r soc02a,fig.align="center",tidy=F,warning=F,message=F,cache=T}
model3 <- lm(call ~ race + sex, data = resume)
model3
# Call rates for a white male
 0.066534 + 0.032130 - 0.009128
```
* Regression Equation: $Y = 0.066 + 0.032*white - 0.009*male + \epsilon$
]
---

# Interpreting Multiple Predictors

.font150[
* .orange[Ceteris Paribus]: _holding everything else constant_

* Let's interpret the coefficient for _white_ in $Y = 0.066 + 0.032 \times white - 0.009 \times male + \epsilon$

* We say: "_all else equal, having a white-sounding name increases the change of getting a job call in about 3%. Since candidates with black-sounding names a job call rate of about 6%, candidates with white-sounding names are 50% more likely to get a call_"
]

---

# Adjusted R-Squared

.font140[
* When we have more than one independent variable, we need to modify the $R^2$ formula to account for those additional variables

* $R^2$ measures the overall impact of _all_ variables at once, but some might just add noise to the model

* Every predictor added to a model increases $R^2$ and never decreases it

* Adjusted $R^2$ compensates for the addition of variables and only increases _if the new term enhances the model_

* It is usually lower than regular $R^2$ but not much
]
---

# Adjusted R-Squared

.font110[
```{r arsq,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=18}
summary(model.cm)
```
]

---

# Merging Data Sets in R

.font110[
* We often have to merge data from different sources

* If the data refer to the same units/indiviuals, .orange[it is essential that both data sets have the same identification variable]

* Use `merge(x, y, by =)` function
]

```{r merge,fig.align="center",tidy=F,warning=F,message=F,cache=T}
pres2008 <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/prediction/pres08.csv")
names(pres2008)
intrade08 <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/prediction/intrade08.csv")
names(intrade08)
{{pres.merged <- merge(pres2008, intrade08, by = "state")}}
```
---

# Merging Data Sets in R

.font110[
```{r merge01,fig.align="center",tidy=F,warning=F,message=F,cache=F}
names(pres.merged)
head(pres.merged)
```
]
---

class: inverse, center, middle

# Questions?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html>  
---

# Homework

.font150[
* `swirl()` PREDICTION2
]
---

class: inverse, center, middle

# See you!

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html>  