---
title: "Week 07 - Prediction"
subtitle: "Model Fit, LM for Randomised Experiments, Multiple Predictors, Merging Data Sets in R<html><div style='float:left'></div><hr color='#EB811B' size=1px width=800px></html>"
author: "Danilo Freire"
date: "8th March 2019"
output: 
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    nature: 
      beforeInit: "macros.js"
      highlightStyle: github
      ratio: 4:3
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
r <- getOption("repos")
r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
options(repos = r)
set.seed(12345)
```

<style>

.remark-slide-number {
  position: inherit;
}

.remark-slide-number .progress-bar-container {
  position: absolute;
  bottom: 0;
  height: 6px;
  display: block;
  left: 0;
  right: 0;
}

.remark-slide-number .progress-bar {
  height: 100%;
  background-color: #EB811B;
}

.orange {
  color: #EB811B;
}
</style>

# Today's Agenda

.font150[
* Review of least squares

* How to assess model fit

* Merging data sets with `merge()`

* Analyse randomised experiments with `lm()`
]
---

# Review

.font150[
* Linear regression minimised the squared distance of the points

* Regression line always goes through the point of averages $\hat{X},\hat{Y}$

* .orange[Regression equation]: $Y = \alpha + \beta X + \epsilon$
  - $\alpha$ is the intercept (value of $Y$ when $X = 0$)
  - $\beta$ is the slope (increase/decrease in $Y$ when $X$ varies by one unit)
  - $\epsilon$ is the error term (difference from points to fitted line; mean zero)
]
---

# Linear Models

.font120[
```{r lm01,fig.align="center",tidy=F,warning=F,message=F,cache=T}
bivariate <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/prediction/bivariate_data.csv")
bivariate <- subset(bivariate, Year == 2010)
model1 <- lm(Child.Mortality ~ log(GDP), data = bivariate)
model1
```
* $Y = 276.58 - 26.13X + \epsilon$
]
---

# R-Squared

.font140[
* How well does our regression line fit the data?

* How well does the model predict the outcome?

* $R^2$, _coefficient of determination_

$$R^2 = 1 - \frac{\textsf{SSR}}{\textsf{Total sum of squares (TSS)}}  \ = \ 1 - \frac{\sum_{i=1}^n \hat\epsilon_i^2}{\sum_{i=1}^n (Y_i - \overline{Y})^2}$$
* $R^2$ is also defined as the _explained variance_ in Y

* Goes from 0 to 1

* We can use `summary(model)` to see it
]
---

# Example

.font110[
```{r rsq2,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=18}
summary(model1)
```
]
---

# Check the Residuals

.font110[
```{r res,fig.align="center",tidy=F,warning=F,message=F,cache=T,fig.height=6.5}
plot(fitted(model1), resid(model1), main = "Residuals",
     xlab = "Fitted values", ylab = "Residuals")
```
]
---

# Check the Residuals

.font120[
```{r res02,fig.align="center",tidy=F,warning=F,message=F,cache=T}
# Which countries are the most relevant outliers?
bivariate$Entity[resid(model1) == max(resid(model1))]

# Which country is the closest to the predicted line?
bivariate$Entity[resid(model1) == min(resid(model1))]
```
]
---

# Caveats

.font150[
* $R^2$ cannot be used to compare models estimated with different samples

* $R^2$ .orange[does not] indicate whether:
  - _the independent variable causes the dependent variable_
  - the model is correctly specified 
  - the model suffers from ommitted variable bias
  - there are enough data points to make a valid conclusion
]
---

# Linear Models and RCTs

.font150[
* .orange[When the data come from a randomised experiment,] model parameters have a causal interpretation

* Treatment status as the independent variable (0 or 1)
  - 0 indicates control group
  - 1 indicates treatment group
]
--
.font150[
* $Y = \alpha + \beta * treatment + \epsilon$
]
--
.font150[
* What is the interpretation of $\alpha$ here?

* What is the interpretation of $\beta$ here?
]
---

# Women as Policy Makers

.font130[
* Do women promote different policies than men?

* Observational studies: compare policies adopted by female politicians with those adopted by male politicians

* Randomised natural experiment:
  - one third of village council heads reserved for women
  - assigned at the level of Gram Panchayat (GP) since mid-1990s
  - each GP has multiple villages

* Hypothesis: female politicians represent the interests of female voters

* Female voters complain about drinking water while male voters complain about irrigation
]
---

# Data

| Name         | Description                                                                                                                |
| :----------- | :------------------------------------------------------------------------------------------------------------------------- |
| `GP`         | An identifier for the Gram Panchayat (GP)                                                                                  |
| `village`    | identifier for each village                                                                                                |
| `reserved`   | binary variable indicating whether the GP was reserved for women leaders or not                                            |
| `female`     | binary variable indicating whether the GP had a female leader or not                                                       |
| `irrigation` | variable measuring the number of new or repaired irrigation facilities in the village since the reserve policy started     |
| `water`      | variable measuring the number of new or repaired drinking-water facilities in the village since the reserve policy started |

```{r data,fig.align="center",tidy=F,warning=F,message=F,cache=T}
women <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/prediction/women.csv")
names(women)
```
---

# Models

.font110[
* Does the reservation policy increase female politicians?

```{r m1,fig.align="center",tidy=F,warning=F,message=F,cache=T}
tapply(women$female, women$reserved, mean)
```

* Does it change the policy outcomes?
```{r m2,fig.align="center",tidy=F,warning=F,message=F,cache=T}
# drinking-water facilities
tapply(women$water, women$reserved, mean)

# irrigation facilities
tapply(women$irrigation, women$reserved, mean)
```
]
---

# Slope Coefficient = Difference-in-Means

.font120[
```{r women03,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=2}
tapply(women$water, women$reserved, mean)
```
```{r women04,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=1}
mean(women$water[women$reserved == 1]) - mean(women$water[women$reserved == 0])
```
```{r women05,fig.align="center",tidy=F,warning=F,message=F,cache=T,highlight.output=7}
lm(water ~ reserved, data = women)
```
]
---

# Linear Regression with Multiple Predictors

.font130[
* The Model:

* $Y = \alpha + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_n X_n + \epsilon$

* Sum of squared residuals (SSR):

* $\textsf{SSR} = \sum_{i=1}^n \hat\epsilon_i^2 \ = \ \sum_{i=1}^n (Y_i - \hat\alpha - \hat\beta_1 X_{i1} - \hat\beta_2 X_{i2} - \cdots - \hat\beta_n X_{in})^2$

* In R: `lm(Y ~ X1 + X2 + ... + Xn, data = yourdataset)` 
]
---

# Social Pressure Experiment Revisited

.font110[
```{r soc01,fig.align="center",tidy=F,warning=F,message=F,cache=T}
social <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/causality/social.csv")
levels(social$messages) # base level is `Civic'
fit <- lm(primary2006 ~ messages, data = social)
fit
```
* The `lm()` function automatically creates an indicator variable for each level of a factor variable
]
---

# Social Pressure Experiment Revisited

.font110[
* The predicted values give the average outcome under each condition

```{r soc02,fig.align="center",tidy=F,warning=F,message=F,cache=T}
unique(social$messages)
predict(fit, newdata = data.frame(messages = unique(social$messages)))
tapply(social$primary2006, social$messages, mean)
```
]
---

# Merging Data Sets in R

.font110[
* We often have to merge data from different sources

* If the data refer to the same units/indiviuals, .orange[it is essential that both data sets have the same identification variable]

* Use `merge(x, y, by =)` function
]

```{r merge,fig.align="center",tidy=F,warning=F,message=F,cache=T}
pres2008 <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/prediction/pres08.csv")
names(pres2008)
intrade08 <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/prediction/intrade08.csv")
names(intrade08)
{{pres.merged <- merge(pres2008, intrade08, by = "state")}}
```
---

# Merging Data Sets in R

.font110[
```{r merge01,fig.align="center",tidy=F,warning=F,message=F,cache=F}
names(pres.merged)
head(pres.merged)
```
]
---

class: inverse, center, middle

# Questions?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html>  
---

# Homework

.font150[
* `swirl()` PREDICTION1
]
---

class: inverse, center, middle

# See you on Friday!

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html>  