---
title: "Week 03 - Observational Studies"
subtitle: "Descriptive Statistics for a Single Variable <html><div style='float:left'></div><hr color='#EB811B' size=1px width=800px></html>"
author: "Danilo Freire"
date: "8th February 2019"
output: 
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    nature: 
      beforeInit: "macros.js"
      highlightStyle: github
      ratio: 4:3
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
r <- getOption("repos")
r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
options(repos = r)
```

<style>

.remark-slide-number {
  position: inherit;
}

.remark-slide-number .progress-bar-container {
  position: absolute;
  bottom: 0;
  height: 6px;
  display: block;
  left: 0;
  right: 0;
}

.remark-slide-number .progress-bar {
  height: 100%;
  background-color: #EB811B;
}

.orange {
  color: #EB811B;
}
</style>

# Today's Agenda

.font150[
* Quantiles

* Standard deviation

* Barplots

* Leader assassination DiD 

* Interval and external validity
]

---

# Quantiles

.font150[
* Sometimes it is useful to look at the distribution of a given variable

* We can split a variable in many ways:
	- Quartiles
	- Quantiles
	- Percentiles

* Which quantile is the median?
]
---

# Quantiles

.font150[
* What is the median of {2, 5, 6, 10}?

* What is the median of {1, 2, 3, 4, 20}?

* Interquartile range (IQR): the difference between the 75th and the 25th percentile    
]

---

# Standard Deviation

.font150[
* Average distance of each data point to the mean

* $SD = (\sqrt{\frac{1}{n} \sum^{N}_{i = 1} (x_{i} - \bar{x})^{2}})$
]
--

.font150[
* where $\bar{x}$ indicates the sample mean, that is, $\bar{x} = \frac{1}{n} \sum^{N}_{i = 1} x_{i}$, and $n$ is the sample size 

* Almost all data entries are located within 2 or 3 standard deviations from the mean
]
---

# R Examples

```{r tidy=FALSE, echo=FALSE}
leaders <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/causality/leaders.csv")
```

```{r tidy=FALSE}
median(leaders$age)
IQR(leaders$age)
quantile(leaders$age)
quantile(leaders$age, probs = seq(0, 1, by = 0.1)) # deciles
quantile(leaders$age, probs = c(.34, .55, .93)) # 34th, 55th, 93th percentiles
```
---

# R Examples

```{r}
mean(leaders$age)
sd(leaders$age)
summary(leaders$age)
```
---

class: inverse, center, middle

# Questions?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 

---

# Leader Assassination DD

.font150[
* **Question 5**

* Does successful leader assassination cause democratization? Does successful leader assassination lead countries to war?  Answer these questions by analyzing the data.  Be sure to state your assumptions and provide a brief interpretation of the results.]
---

# Leader Assassination DD

```{r}
str(leaders)

# create success variable
leaders$success <- ifelse(leaders$result == "dies between a day and a week" |
                            leaders$result == "dies between a week and a month" |
                            leaders$result == "dies within a day after the attack" |
                            leaders$result == "dies, timing unknown",1, 0)
```
---

# Leader Assassination DD

```{r}
# polity score before and after assassination attempt
diff.pol.succ <- mean(leaders$polityafter[leaders$success == 1]) - 
                   mean(leaders$politybefore[leaders$success == 1]) # successful 

diff.pol.succ

diff.pol.unsucc <- mean(leaders$polityafter[leaders$success == 0]) - 
                     mean(leaders$politybefore[leaders$success == 0]) # unsuccessful

diff.pol.unsucc

## difference in differences
diff.pol.succ - diff.pol.unsucc
```
---

# Leader Assassination DD

```{r tidy=FALSE}
# create variable for warbefore warafter
leaders$warbefore <- ifelse(leaders$interwarbefore == 1 |
                              leaders$civilwarbefore == 1, 1, 0)
leaders$warafter <- ifelse(leaders$interwarafter == 1 |
                             leaders$civilwarafter == 1, 1, 0)

## compare war before to war after among successful and unsuccessful
diff.war.succ <- mean(leaders$warafter[leaders$success == 1]) - 
                   mean(leaders$warbefore[leaders$success == 1])

diff.war.unsucc <- mean(leaders$warafter[leaders$success == 0]) - 
                     mean(leaders$warbefore[leaders$success == 0])

diff.war.succ - diff.war.unsucc # difference in differences
```

.font130[
Using the difference-in-difference approach, we find very little difference in the contries' polity score and in the proportion of countries engaged in war. Leader assassination does not seem to cause countries to democratise or engage in war.]
---

# Homework

.font150[
* `CAUSALITY02`
]

---

class: inverse, center, middle

# See you on Friday!

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 
