---
title: "Week 04 - Measurement"
subtitle: "Missing Observations and Data Visualisation <html><div style='float:left'></div><hr color='#EB811B' size=1px width=800px></html>"
author: "Danilo Freire"
date: "13th February 2019"
output: 
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    nature: 
      beforeInit: "macros.js"
      highlightStyle: github
      ratio: 4:3
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
r <- getOption("repos")
r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
options(repos = r)
```

<style>

.remark-slide-number {
  position: inherit;
}

.remark-slide-number .progress-bar-container {
  position: absolute;
  bottom: 0;
  height: 6px;
  display: block;
  left: 0;
  right: 0;
}

.remark-slide-number .progress-bar {
  height: 100%;
  background-color: #EB811B;
}

.orange {
  color: #EB811B;
}
</style>

# Today's Agenda

.font150[
* Tables: `table()` and `prop.table()` 

* Missing data in R: `is.na()` and `na.omit()`

* Visualising data: `barplot()` and `hist()`
]

---

# Tables in R

.font130[
* We have learned how to use the `table()` command in R

* Example:
]

.code120[
```{r tidy=FALSE}
afghan <- read.csv("https://raw.githubusercontent.com/pols1600/pols1600.github.io/master/datasets/measurement/afghan.csv")
names(afghan)
table(ISAF = afghan$violent.exp.ISAF,
      Taliban = afghan$violent.exp.taliban)
```
]
---

# Tables in R: prop.table()

.font130[
* We can also use `prop.table()` to see the proportion of cases in each cell

* We have to include `table()` within parentheses too:
]

.code120[
```{r tidy=FALSE}
prop.table(table(ISAF = afghan$violent.exp.ISAF,
                 Taliban = afghan$violent.exp.taliban))
```
]
---
# Tables in R

.font130[
* Since we're already using nested functions, we can also use `round()` to round the values in each cell

* Notice the `, 2` in the code below. It indicates that we will round the numbers up to two significant digits
]

.code120[
```{r tidy=FALSE}
round(prop.table(table(ISAF = afghan$violent.exp.ISAF,
                       Taliban = afghan$violent.exp.taliban)), 2)
```
]

.font150[
* The table is now much easier to read and it conveys the same information
]
---

# Tables in R: prop.table()

.code120[
```{r tidy=FALSE}
round(prop.table(table(ISAF = afghan$violent.exp.ISAF,
                       Taliban = afghan$violent.exp.taliban)), 2)
```
]

.font150[
* What's the percentage of people that have been victimised by the Taliban?

* And by both the Taliban and ISAF?
]
---

# Tables in R: prop.table()

.font150[
* Another example, now with 3 significant digits and multiplied by 100
]

.code120[
```{r tidy=FALSE}
round(prop.table(table(Employed = afghan$employed,
                       Income = afghan$income)), 3) * 100
```
]
---

# Missing Data

.font150[
* Not all individuals answer to surveys

* Two types of non-response:
  - Individual non-response
  - Item non-response
  
* Both tend to bias the results

* So it is very important that we know where (and think about why) we see gaps in our data
]
---

# Missing Data

.font150[
* R has a special code for missing data, `NA`

* Since `NA` is only used for missing observations, we can count their numbers with `is.na()`
]

---

# Missing Data

.code120[
```{r}
head(afghan$income, n = 10)
sum(is.na(afghan$income)) # number of missings
round(mean(is.na(afghan$income)), 2) # proportion of missings
```
]
---

# Missing Data

.font150[
* Some R function don't work if there's missing data

* We add `na.rm = TRUE` to the code
]

.code120[
```{r}
# Victims of Taliban violence
sum(is.na(afghan$violent.exp.taliban))
mean(afghan$violent.exp.taliban)
round(mean(afghan$violent.exp.taliban, na.rm = TRUE), 2)
```
]

.font150[
* Why do you think we have missing data here?
]

---

# Missing Data

.font150[
* We can create a new variable or data set without missing observations

* Use the `na.omit()` command
]

.code150[
```{r}
length(afghan$violent.exp.taliban)
taliban.no.missing <- na.omit(afghan$violent.exp.taliban)
length(taliban.no.missing)
length(afghan$violent.exp.taliban) - length(taliban.no.missing)
```
]

---

# Missing Data

.font150[
* You can also visualise the number of missing observations with `summary()`
]

.code150[
```{r}
summary(afghan$violent.exp.taliban)
sum(is.na(afghan$violent.exp.taliban))
```
]
---

class: inverse, center, middle

# Data Visualisation

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 
---

# Bar Plots

.font120[
* Bar plots are used to visualise .orange[factor/character variables]

* Proportion of observations in each category as the height of each bar

* Options:

  - `main = "Title"`
  - `xlab = "X label"`
  - `ylab = "Y label"`
  - `xlim = c(number, number)` limits for the x variable  
  - `ylim = c(number, number)` limits for the y variable  
  - `names.arg = c("Bars labels")` - in the same order of the variable
  - `horiz = TRUE` for horizontal plots
  - `cols = "colour name"` bar colour (see: )
  
* You can use `barplot()` with `prop.table()` instead of pie charts
]
---

# Bar Plots

```{r}
employed.ptable <- prop.table(table(afghan$employed))
employed.ptable
```

```{r tidy=FALSE, eval=FALSE}
employed.ptable <- prop.table(table(afghan$employed))
barplot(employed.ptable,
        names.arg = c("Unemployed", "Employed"), 
        main = "Proportion of Employed Afghanis",
        xlab = "Employment",
        ylab = "Proportion",
        ylim = c(0, 0.6))
```
---

# Bar Plots

```{r tidy=FALSE, echo=FALSE, fig.align='center'}
barplot(employed.ptable,
        names.arg = c("Unemployed", "Employed"), # 0 and 1, respectively
        main = "Proportion of Employed Afghanis",
        xlab = "Employment",
        ylab = "Proportion",
        ylim = c(0, 0.6))
```

---

# Bar Plots

```{r tidy=FALSE, eval=FALSE}
barplot(employed.ptable,
        names.arg = c("Unemployed", "Employed"), # 0 and 1, respectively
        main = "Proportion of Employed Afghanis",
        ylab = "Employment", # change the axes
        xlab = "Proportion", 
        xlim = c(0, 0.7), # now it's xlim 
        horiz = TRUE,     # because the plot is horizontal
        col = "brown")
```
---

# Bar Plots

```{r tidy=FALSE, echo=FALSE, fig.align='center'}
barplot(employed.ptable,
        names.arg = c("Unemployed", "Employed"), # 0 and 1, respectively
        main = "Proportion of Employed Afghanis",
        ylab = "Employment",
        xlab = "Proportion",
        xlim = c(0, 0.6),
        horiz = TRUE,
        col = "brown")
```

---

# Histograms

.font120[
* We use histograms to display the distribution of a .orange[numeric variable]

* They are similar to bar plots

* Numeric variables are _binned_ into groups

* Histograms shows the density of each bin

* Important: Height is share of observations in bin divided by bin size

* We care less about the density of each bin than about the distribution of the variable as a whole

* Area of each bar is the share of observations that fall into that bin

* Area of all bins sum to one
]
---

# Histograms

.font120[
* Many options are similar to those of `barplot()`: `main`, `xlab`, `ylim`, `col`

* We can also add `freq = FALSE` to show the density of each histograms

* `breaks =` changes the size of the bins

* Densities are useful to compare different distributions

* .orange[Densities are not percentages]: "percentage per horizontal unit"
]
---

# Histograms

```{r tidy=FALSE, eval=FALSE}
# For colours, see: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
hist(afghan$age,
     main = "Histogram - Age",
     xlab = "Age",
     xlim = c(0, 0.04),
     freq = FALSE,
     col = "darkorange2") 
```
---

# Histograms

```{r tidy=FALSE, echo=FALSE, fig.align='center'}
hist(afghan$age,
     main = "Histogram - Age",
     xlab = "Age",
     ylim = c(0, 0.04),
     freq = FALSE,
     col = "darkorange2") 
```
---

# Histograms

.font150[
* We can also add text and fitted lines to all R plots

* Use `text()` and `abline()` after `hist()`
]

```{r tidy=FALSE, eval=FALSE}
hist(afghan$age,
     main = "Histogram - Age",
     xlab = "Age",
     ylim = c(0, 0.04),
     freq = FALSE,
     col = "darkorange2") 
## add a text label at (x, y) = (35, 0.35)
text(x = 35, y = 0.035, "median")
## add a vertical line representing median
abline(v = median(afghan$age))
```
---

# Histograms

```{r tidy=FALSE, echo=FALSE, fig.align='center'}
hist(afghan$age,
     main = "Histogram - Age",
     xlab = "Age",
     ylim = c(0, 0.04),
     freq = FALSE,
     col = "darkorange2") 
## add a text label at (x, y) = (35, 0.35)
text(x = 35, y = 0.035, "median")
## add a vertical line representing median
abline(v = median(afghan$age))
```
---

class: inverse, center, middle

# Questions?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 
---

# Homework

.font150[
* `MEASUREMENT01`
]

---

class: inverse, center, middle

# See you on Friday!

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=720px></html> 
